---
title: "04_ASE_lactate_analysis"
author: "Kevin Buell, Will Parker"
date: "2024-10-27"
output: html_document
---
# Required packages
```{r}
library(here)
library(tidyverse)
library(gtsummary)
library(arrow)
library(readr)
library(glue)
library(networkD3)
```

# Load in sepsis study cohort CLIF tables
```{r}
# Load the configuration utility
source(here("utils/config.R"))

site_name <- config$site_name
tables_path <- file.path(config$tables_path, "sepsis_study_cohort")
file_type <- config$file_type
output_folder <- here::here(paste0("result_", site_name)) 

start_date <- as.Date(config$date_range$start)
end_date <- as.Date(config$date_range$end)

# Get a list of all parquet files in the folder
parquet_files <- list.files(tables_path, pattern = "*.parquet", full.names = TRUE)

# Function to load parquet file and assign to a variable with the name based on the file name
load_parquet_file <- function(file_path) {
  # Create a variable name from the file name without extension
  var_name <- tools::file_path_sans_ext(basename(file_path))
  
  # Load the parquet file and assign it to the global environment without previewing it
  assign(var_name, read_parquet(file_path), envir = .GlobalEnv)
  invisible(NULL) # Suppress any output from the function
}

# Use purrr::walk instead of lapply to suppress unnecessary output
purrr::walk(parquet_files, load_parquet_file)

glue("CLIF tables for {site_name} loaded.")
```

```{r}
sepsis_case_summary <- sepsis_case_summary %>%
    mutate(
    lactate_only_dx = ifelse(
      !is.na(lactate) & is.na(thrombocytopenia) & is.na(aki) &
      is.na(`invasive mechanical ventilation`) & is.na(vasopressor) &
      is.na(hyperbilirubinemia), 
      1, 0
    ),
    # Earliest time with lactate
    earliest_time_with_lactate = pmin(thrombocytopenia, aki, `invasive mechanical ventilation`,
                                      lactate, vasopressor, hyperbilirubinemia, na.rm = TRUE),
    
    # Earliest time without lactate
    earliest_time_without_lactate = pmin(thrombocytopenia, aki, `invasive mechanical ventilation`,
                                         vasopressor, hyperbilirubinemia, na.rm = TRUE),
     first_organ_dysfunction_with_lactate = case_when(
      thrombocytopenia == earliest_time_with_lactate ~ "thrombocytopenia",
      aki == earliest_time_with_lactate ~ "aki",
      `invasive mechanical ventilation` == earliest_time_with_lactate ~ "invasive mechanical ventilation",
      lactate == earliest_time_with_lactate ~ "lactate",
      vasopressor == earliest_time_with_lactate ~ "vasopressor",
      hyperbilirubinemia == earliest_time_with_lactate ~ "hyperbilirubinemia",
      TRUE ~ NA_character_
    ),
    first_organ_dysfunction_without_lactate = case_when(
      thrombocytopenia == earliest_time_without_lactate ~ "thrombocytopenia",
      aki == earliest_time_without_lactate ~ "aki",
      `invasive mechanical ventilation` == earliest_time_without_lactate ~ "invasive mechanical ventilation",
      vasopressor == earliest_time_without_lactate ~ "vasopressor",
      hyperbilirubinemia == earliest_time_without_lactate ~ "hyperbilirubinemia",
      TRUE ~ NA_character_
    ),
    dx_time_gained_with_lactate = as.numeric(difftime(earliest_time_without_lactate, earliest_time_with_lactate, units = "hours"))
  )
```

```{r}
sepsis_case_summary <- sepsis_case_summary %>%
  left_join(clif_hospitalization_cohort %>% select(hospitalization_id, discharge_category, 
                                                   admission_dttm, admission_type_name, admission_type_category)) %>%
    mutate(mortality = ifelse(discharge_category %in% c("Expired", "Hospice"), "died", "survived"),
        time_since_admission_w_lactate = difftime(earliest_time_with_lactate, admission_dttm, units = "days"),
         onset_w_lactate = ifelse(time_since_admission_w_lactate <= 2, "community", "hospital-acquired"),
        time_since_admission_without_lactate = difftime(earliest_time_without_lactate, admission_dttm, units = "days"),
        onset_without_lactate = ifelse(time_since_admission_without_lactate <= 2, "community", "hospital-acquired"))
```

# Time to sepsis onset
```{r}
sepsis_case_summary %>%
  ggplot(aes(x = time_since_admission_w_lactate, fill = onset_w_lactate)) +
  geom_histogram(breaks = seq(0, 14)) + labs(x = "Days since admission", color = "Onset category", title = "Time to sepsis diagnosis with lactate in the definition") 
```

```{r}
sepsis_case_summary %>%
  ggplot(aes(x = time_since_admission_without_lactate, fill = onset_without_lactate)) +
  geom_histogram(breaks = seq(0, 14)) + labs(x = "Days since admission", color = "Onset category", title = "Time to sepsis diagnosis without lactate in the definition") 
```
```{r}

# Step 1: Prepare the nodes and links for the Sankey plot

# Define the nodes
nodes <- data.frame(
  name = c("Community-Onset with Lactate", "Hospital-Onset with Lactate", 
           "Community-Onset without Lactate", "Hospital-Onset without Lactate", 
           "Not Sepsis without Lactate")
)

# Create a mapping of categories in the original data to node indices
node_map <- c(
  "community_community" = 0,
  "community_hospital-acquired" = 1,
  "community_NA" = 4,
  "hospital-acquired_hospital-acquired" = 3,
  "hospital-acquired_NA" = 4
)

# Step 2: Prepare the links based on the data
links <- sepsis_case_summary %>%
  group_by(onset_w_lactate, onset_without_lactate) %>%
  summarize(n = n()) %>%
  mutate(
    source = case_when(
      onset_w_lactate == "community" & onset_without_lactate == "community" ~ 0,
      onset_w_lactate == "community" & onset_without_lactate == "hospital-acquired" ~ 0,
      onset_w_lactate == "community" & is.na(onset_without_lactate) ~ 0,
      onset_w_lactate == "hospital-acquired" & onset_without_lactate == "hospital-acquired" ~ 1,
      onset_w_lactate == "hospital-acquired" & is.na(onset_without_lactate) ~ 1
    ),
    target = case_when(
      onset_w_lactate == "community" & onset_without_lactate == "community" ~ 2,
      onset_w_lactate == "community" & onset_without_lactate == "hospital-acquired" ~ 3,
      onset_w_lactate == "community" & is.na(onset_without_lactate) ~ 4,
      onset_w_lactate == "hospital-acquired" & onset_without_lactate == "hospital-acquired" ~ 3,
      onset_w_lactate == "hospital-acquired" & is.na(onset_without_lactate) ~ 4
    )
  ) %>%
  select(source, target, value = n)

# Step 3: Create the Sankey plot
sankeyNetwork(Links = links, Nodes = nodes, Source = "source", Target = "target", 
              Value = "value", NodeID = "name", units = "Cases", 
              fontSize = 12, nodeWidth = 30)
```


```{r}
onset_summary <- sepsis_case_summary %>%
  group_by(onset_w_lactate, onset_without_lactate) %>%
  summarize(n = n())

onset_summary %>%
  mutate(onset_without_lactate = ifelse(is.na(onset_without_lactate), "Not sepsis without lactate", onset_without_lactate))

write_dataframe_to_csv <- function(dataframe, output_folder = here::here(paste0("result_", site_name))) {
  # Get the name of the dataframe as a string
  dataframe_name <- deparse(substitute(dataframe))
  
  # Define the file path using the dataframe name
  file_path <- file.path(output_folder, paste0(dataframe_name, ".csv"))
  
  # Write the dataframe to CSV
  write_csv(dataframe, file_path)
  
  # Print a message indicating where the file was saved
  message(glue::glue("Dataframe '{dataframe_name}' has been written to '{file_path}'"))
}

write_dataframe_to_csv(onset_summary)
```
#Calculate organ dysfunction incidence
```{r}
sepsis_ids <- sepsis_case_summary %>% pull(hospitalization_id)

shock_ever <- long_vasoactives %>% pull(hospitalization_id) %>% unique()
imv_ever <- long_IMV %>% pull(hospitalization_id) %>% unique()
aki_ever <- long_sepsis_labs %>% filter(aki ==1) %>% pull(hospitalization_id) %>% unique()
hyperbili_ever <- long_sepsis_labs %>% filter(hyperbilirubinemia ==1) %>% pull(hospitalization_id) %>% unique()
thrombocytopenia_ever <- long_sepsis_labs %>% filter(thrombocytopenia ==1) %>% pull(hospitalization_id) %>% unique()
lactate_ever <- long_sepsis_labs %>% filter(lab_category == "lactate" & lab_value_numeric > 2 ) %>% pull(hospitalization_id) %>% unique()


# Add binary indicator variables to sepsis_case_summary
sepsis_case_summary <- sepsis_case_summary %>%
  mutate(
    shock_ever = if_else(hospitalization_id %in% shock_ever, 1, 0),
    imv_ever = if_else(hospitalization_id %in% imv_ever, 1, 0),
    aki_ever = if_else(hospitalization_id %in% aki_ever, 1, 0),
    hyperbili_ever = if_else(hospitalization_id %in% hyperbili_ever, 1, 0),
    thrombocytopenia_ever = if_else(hospitalization_id %in% thrombocytopenia_ever, 1, 0),
    lactate_ever = if_else(hospitalization_id %in% lactate_ever, 1, 0)
  )

```

```{r}
organ_dysfunction_counts_all <- sepsis_case_summary %>%
  ungroup() %>%
  select(ends_with("_ever")) %>%
  summarise(
    shock_ever_count = sum(shock_ever),
    imv_ever_count = sum(imv_ever),
    aki_ever_count = sum(aki_ever),
    hyperbili_ever_count = sum(hyperbili_ever),
    thrombocytopenia_ever_count = sum(thrombocytopenia_ever),
    lactate_ever_count = sum(lactate_ever)
  ) %>%
  mutate(site_name = site_name)

write_dataframe_to_csv(organ_dysfunction_counts_all)
```

```{r}
organ_dysfunction_counts_no_lactate <- sepsis_case_summary %>%
  ungroup() %>%
  filter(lactate_only_dx ==0) %>%
  select(ends_with("_ever")) %>%
  summarise(
    shock_ever_count = sum(shock_ever),
    imv_ever_count = sum(imv_ever),
    aki_ever_count = sum(aki_ever),
    hyperbili_ever_count = sum(hyperbili_ever),
    thrombocytopenia_ever_count = sum(thrombocytopenia_ever),
    lactate_ever_count = sum(lactate_ever)
  ) %>%
  mutate(site_name = site_name)

write_dataframe_to_csv(organ_dysfunction_counts_no_lactate)
```

#Calculate overall sepsis incidence, with and without inclusion of lactate 
```{r}
#all patients who meet part B sepsis criteria (lactate, vasopressos, IMV, creatinine, bili, platelets) including use of lactate 
sepsis_n_including_lactate<- sepsis_case_summary %>%
  pull(hospitalization_id) %>%
  unique() %>%
  length()

#patients who meet part B sepsis criteria because of addition of lactate 
sepsis_n_becauseof_lactate <- sepsis_case_summary %>%
  
  filter(lactate_only_dx == 1) %>%
  pull(hospitalization_id) %>% 
  unique() %>%
  length()

#patients who meet part B sepsis criteria (vasopressos, IMV, creatinine, bili, platelets) excluding use of lactate 
sepsis_n_excluding_lactate <- sepsis_n_including_lactate - sepsis_n_becauseof_lactate

#all hospitalizations in cohort
total_hospitalizations <- clif_hospitalization_cohort %>% nrow()

#calculate sepsis incidence
sepsis_incidence <- tibble(
  site = as.factor(site_name),  
  sepsis_n_including_lactate,  
  sepsis_n_becauseof_lactate,  
  sepsis_n_excluding_lactate,
  total_hospitalizations) %>%  
mutate(sepsis_incidence_including_lactate = round(100 * sepsis_n_including_lactate / total_hospitalizations, digits = 2),
       sepsis_incidence_excluding_lactate = round(100 * sepsis_n_excluding_lactate / total_hospitalizations, digits = 2))


#check sepsis incidence is reasonable
glue(
"For {site_name}, the sepsis incidence was {round(100 * sepsis_n_including_lactate / total_hospitalizations, digits = 2)}% for adult patients admitted from {start_date} to {end_date}. The sepsis incidence without lactate was only {round(100 * sepsis_n_excluding_lactate / total_hospitalizations, digits = 2)}%."
)

```

```{r}
write_dataframe_to_csv(sepsis_incidence)
```

```{r}
sepsis_summary_table <- sepsis_case_summary %>%
  ungroup() %>%
  mutate(
         sepsis_with_lactate = !is.na(earliest_time_with_lactate),
    sepsis_without_lactate = !is.na(earliest_time_without_lactate)) %>%
  summarise(
    sepsis_cases_with_lactate = sum(sepsis_with_lactate, na.rm = TRUE),
    sepsis_cases_without_lactate = sum(sepsis_without_lactate, na.rm = TRUE),

    mortality_with_lactate = sum(sepsis_with_lactate & mortality == "died", na.rm = TRUE),
    mortality_without_lactate = sum(sepsis_without_lactate & mortality == "died", na.rm = TRUE),
  ) %>%
  mutate(healthsystem = site_name,
         total_hospitalizations = total_hospitalizations,
         incidence_with_lactate = round((sepsis_cases_with_lactate / total_hospitalizations) * 100, 2),
    incidence_without_lactate = round((sepsis_cases_without_lactate / total_hospitalizations) * 100, 2),
        mortality_rate_with_lactate = round((mortality_with_lactate / sepsis_cases_with_lactate) * 100, 2),
    mortality_rate_without_lactate = round((mortality_without_lactate / sepsis_cases_without_lactate) * 100, 2))


sepsis_summary_table <- sepsis_summary_table %>% select(healthsystem, total_hospitalizations, incidence_with_lactate, incidence_without_lactate, mortality_rate_with_lactate, mortality_rate_without_lactate)
```


#Calculate sepsis incidence trends over time, with and without inclusion of lactate 
```{r}
#all patients who meet part B sepsis criteria (lactate, vasopressos, IMV, creatinine, bili, platelets) including use of lactate 
sepsis_cases_time_withlactate_time <- sepsis_case_summary %>%
  select(hospitalization_id, earliest_time_with_lactate) %>%
  mutate(year = year(earliest_time_with_lactate),
         month = month(earliest_time_with_lactate)) %>% 
  group_by(year, month) %>% 
  summarise(sepsis_cases_time_includinglactate = n_distinct(hospitalization_id))

#patients who meet part B sepsis criteria because of addition of lactate 
sepsis_n_becauseof_lactate_time <- sepsis_case_summary %>%
  filter(lactate_only_dx == 1) %>%
  select(hospitalization_id, earliest_time_with_lactate) %>%
  mutate(year = year(earliest_time_with_lactate),
         month = month(earliest_time_with_lactate)) %>%
  group_by(year, month) %>% 
  summarise(sepsis_cases_becauseof_lactate_time = n_distinct(hospitalization_id)) 

#patients who meet part B sepsis criteria (vasopressos, IMV, creatinine, bili, platelets) excluding use of lactate 
sepsis_n_excluding_lactate_time <- sepsis_cases_time_withlactate_time %>% 
  left_join(sepsis_n_becauseof_lactate_time, by = c("year", "month")) %>%
  mutate(sepsis_n_excluding_lactate_time = sepsis_cases_time_includinglactate - sepsis_cases_becauseof_lactate_time)

#all hospitalizations in cohort
hospitalizations_time <- clif_hospitalization_cohort %>% 
  select(hospitalization_id, admission_dttm) %>% 
  mutate(year = year(admission_dttm),
         month = month(admission_dttm)) %>% 
  group_by(year, month) %>%
  summarise(total_hospitalizations_time = n_distinct(hospitalization_id)) 


sepsis_incidence_time <- sepsis_n_excluding_lactate_time %>% 
  left_join(hospitalizations_time, by = c("year", "month")) %>%
  arrange(year, month) %>% 
  mutate(sepsis_incidence_including_lactate = round(100 * sepsis_cases_time_includinglactate / total_hospitalizations_time, digits = 2),
       sepsis_incidence_excluding_lactate = round(100 * sepsis_n_excluding_lactate_time / total_hospitalizations_time, digits = 2)) %>% 
  mutate(site = as.factor(site_name))



#plot incidence of sepsis with and without use of lactate 
  sepsis_incidence_time %>%
  select(year, month, sepsis_incidence_including_lactate, sepsis_incidence_excluding_lactate) %>%
  mutate(year_month = ym(paste(year, month, sep = "-"))) %>%
  pivot_longer(cols = starts_with("sepsis_incidence"), 
               names_to = "lactate_used", 
               values_to = "sepsis_incidence") %>% 
  mutate(lactate_used = recode(lactate_used, 
                               "sepsis_incidence_including_lactate" = "Including Lactate",
                               "sepsis_incidence_excluding_lactate" = "Excluding Lactate")) %>%
  ggplot(aes(x = year_month, y = sepsis_incidence, color = lactate_used)) +
  geom_smooth(se = FALSE, span = 0.4) +  # Smoothed line without confidence interval
  scale_x_date(date_breaks = "1 month", date_labels = "%Y-%m") +  # Show all months on x-axis
  labs(title = "Sepsis Incidence over Time",
       x = "Year-Month",
       y = "Sepsis Incidence (%)",
       color = "Lactate Used") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_folder, "sepsis_incidence.pdf"))
```
```{r}
write_dataframe_to_csv(sepsis_incidence_time)
```

#assessment of lactate ordering trends over time; number of lactate ordered per sepsis patient hospitalization
```{r}
lactate_order_per_patient_full_hospitalization <- sepsis_case_summary %>%
  left_join(clif_labs_cohort %>%
              select(hospitalization_id, lab_result_dttm, lab_category),
            by = "hospitalization_id") %>%  # Join on hospitalization_id
  filter(lab_category == "lactate") %>% 
  group_by(hospitalization_id) %>%  # Group by hospitalization_id
  summarise(lactate_orders_n = n_distinct(lab_result_dttm)) %>%  # Count distinct lactate orders
  ungroup() %>% 
mutate(number_lactate = cut(
    lactate_orders_n, 
    breaks = seq(0, 100, by = 1),  # Create breaks from 0 to 30 by 1
    include.lowest = TRUE,        # Include the lowest bound in the first interval
    right = FALSE,                # Left-closed intervals
    labels = paste0(seq(0, 99, by = 1), "-", seq(1, 100, by = 1))  # Create labels "0-1", "1-2", etc.
  )) %>%
  group_by(number_lactate) %>%
  summarise(patients_n = n_distinct(hospitalization_id)) %>%  # Count distinct patients per group
  ungroup() %>% 
  mutate(
    total = sum(patients_n),
    percent.total = patients_n /sum(patients_n) * 100) %>% 
  mutate(timing_sepsis = as.factor("full_hospitalization")) %>% 
  mutate(site = as.factor(site_name)) #please insert your site name here

lactate_order_per_patient_until_sepsis <- sepsis_case_summary %>%
  left_join(clif_labs_cohort %>%
              select(hospitalization_id, lab_result_dttm, lab_category),
            by = "hospitalization_id")  %>% 
  filter(lab_category == "lactate") %>% 
  filter(lab_result_dttm <= earliest_time_with_lactate) %>% #censor after sepsis diagnosis 
  group_by(hospitalization_id) %>%  # Group by hospitalization_id
  summarise(lactate_orders_n = n_distinct(lab_result_dttm)) %>%  # Count distinct lactate orders
  ungroup() %>% 
mutate(number_lactate = cut(
    lactate_orders_n, 
    breaks = seq(0, 100, by = 1),  # Create breaks from 0 to 30 by 1
    include.lowest = TRUE,        # Include the lowest bound in the first interval
    right = FALSE,                # Left-closed intervals
    labels = paste0(seq(0, 99, by = 1), "-", seq(1, 100, by = 1))  # Create labels "0-1", "1-2", etc.
  )) %>%
  group_by(number_lactate) %>%
  summarise(patients_n = n_distinct(hospitalization_id)) %>%  # Count distinct patients per group
  ungroup() %>% 
  mutate(
    total = sum(patients_n),
    percent.total = patients_n /sum(patients_n) * 100) %>% 
  mutate(timing_sepsis = as.factor("until_sepsis_criteria_met")) %>% 
  mutate(site = as.factor(site_name)) #please insert your site name here

# bind results
lactate_order_per_patient <- rbind(lactate_order_per_patient_until_sepsis, lactate_order_per_patient_full_hospitalization) %>% 
  arrange(number_lactate, timing_sepsis)

#graph it 
# Define the factor levels we are interested in: "1-2", "2-3", ..., "19-20"
target_levels <- paste0(1:19, "-", 2:20)

# Filter the dataset to include only the desired factor levels
lactate_order_per_patient_filtered <- lactate_order_per_patient %>%
  filter(number_lactate %in% target_levels)  # Keep only the desired factor levels

# Plot with ggplot2
ggplot(lactate_order_per_patient_filtered, aes(x = number_lactate, y = percent.total, fill = timing_sepsis)) +
  geom_bar(stat = "identity", position = "dodge") +  # Separate bars for each factor level of timing_sepsis
  labs(title = "Percent of Patients by Number of Lactate Orders",
       x = "Number of Lactate Orders",
       y = "Percent Total",
       fill = "Timing Sepsis") +  # Legend title
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

```{r}
write_dataframe_to_csv(lactate_order_per_patient) 
```

#number lactate orders over time
```{r}
lactate_order_per_patient_full_hospitalization_time <- sepsis_case_summary %>%
  left_join(clif_labs_cohort %>%
              select(hospitalization_id, lab_result_dttm, lab_category),
            by = "hospitalization_id") %>%  # Join on hospitalization_id
  filter(lab_category == "lactate") %>% 
  mutate(year = year(lab_result_dttm),
         month = month(lab_result_dttm)) %>%
  group_by(hospitalization_id, year, month) %>%  # Group by hospitalization_id
  summarise(lactate_orders_n = n_distinct(lab_result_dttm)) %>%  # Count distinct lactate orders
  ungroup() %>% 
mutate(number_lactate = cut(
    lactate_orders_n, 
    breaks = seq(0, 100, by = 1),  # Create breaks from 0 to 30 by 1
    include.lowest = TRUE,        # Include the lowest bound in the first interval
    right = FALSE,                # Left-closed intervals
    labels = paste0(seq(0, 99, by = 1), "-", seq(1, 100, by = 1))  # Create labels "0-1", "1-2", etc.
  )) %>%
  group_by(number_lactate, month, year) %>%
  summarise(patients_n = n_distinct(hospitalization_id)) %>%  # Count distinct patients per group
  ungroup() %>% 
  mutate(
    total = sum(patients_n),
    percent.total = patients_n /sum(patients_n) * 100) %>% #note sum is different because hospitalizations can span over multiple months
  mutate(timing_sepsis = as.factor("full_hospitalization")) %>% 
  mutate(site = as.factor(site_name)) #please insert your site name here

lactate_order_per_patient_until_sepsis_time <- sepsis_case_summary %>%
  left_join(clif_labs_cohort %>%
              select(hospitalization_id, lab_result_dttm, lab_category),
            by = "hospitalization_id")  %>% 
  filter(lab_category == "lactate") %>% 
  mutate(year = year(lab_result_dttm),
         month = month(lab_result_dttm)) %>%
  filter(lab_result_dttm <= earliest_time_with_lactate) %>% #censor after sepsis diagnosis 
  group_by(hospitalization_id, year, month) %>%  # Group by hospitalization_id
  summarise(lactate_orders_n = n_distinct(lab_result_dttm)) %>%  # Count distinct lactate orders
  ungroup() %>% 
mutate(number_lactate = cut(
    lactate_orders_n, 
    breaks = seq(0, 100, by = 1),  # Create breaks from 0 to 30 by 1
    include.lowest = TRUE,        # Include the lowest bound in the first interval
    right = FALSE,                # Left-closed intervals
    labels = paste0(seq(0, 99, by = 1), "-", seq(1, 100, by = 1))  # Create labels "0-1", "1-2", etc.
  )) %>%
  group_by(number_lactate, year, month) %>%
  summarise(patients_n = n_distinct(hospitalization_id)) %>%  # Count distinct patients per group
  ungroup() %>% 
  mutate(
    total = sum(patients_n),
    percent.total = patients_n /sum(patients_n) * 100) %>% 
  mutate(timing_sepsis = as.factor("until_sepsis_criteria_met")) %>% 
  mutate(site = as.factor(site_name)) #please insert your site name here

# bind results
lactate_order_per_patient_time <- rbind(lactate_order_per_patient_until_sepsis_time, lactate_order_per_patient_full_hospitalization_time) %>%
  arrange(number_lactate, year, month, timing_sepsis)

# Create the plot
lactate_order_per_patient_time %>%
  filter(timing_sepsis == "until_sepsis_criteria_met") %>%  # Filter by timing_sepsis
  mutate(year_month = paste(year, month, sep = "-")) %>%  # Create a combined year_month column
  filter(number_lactate %in% paste0(1:9, "-", 2:10)) %>%  # Filter for number_lactate levels from 1-2 to 9-10
  ggplot(aes(x = year_month, y = percent.total, color = number_lactate, group = number_lactate)) +
  geom_smooth(se = FALSE, span = 0.75) +  # Use geom_smooth() to add smoothed lines
  labs(title = "Smoothed Percent Total by Year-Month and Number of Lactate Orders",
       x = "Year-Month",
       y = "Percent Total",
       color = "Number of Lactate Orders") +  # Legend title
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +  # Rotate x-axis labels for readability
  theme(axis.text.x = element_text(hjust = 1))
```

```{r}
write_dataframe_to_csv(lactate_order_per_patient_time)
```

#calculate the proportions of first qualifying organ dysfunctions, with and without lactate criteria
#note the NA in the table is normal, given its the patients with partb lactate only
```{r}
# Calculate proportions for first_organ_dysfunction_without_lactate
without_lactate <- sepsis_case_summary %>%
  ungroup() %>%
  count(first_organ_dysfunction_without_lactate) %>%
  mutate(total_n = sum(n),
         percent.total = round(n / sum(n) * 100, 2),
         category = "without_lactate")

# Calculate proportions for first_organ_dysfunction_with_lactate
with_lactate <- sepsis_case_summary %>%
  ungroup() %>%
  count(first_organ_dysfunction_with_lactate) %>%
  mutate(total_n = sum(n),
         percent.total = round(n / sum(n) * 100, 2),
         category = "with_lactate")

# Combine the two datasets into one for comparison and remove 'lactate' and 'NA'
first_organ_dysfunction_wi_wo_lactacte <- bind_rows(
  without_lactate %>% rename(first_organ_dysfunction = first_organ_dysfunction_without_lactate),
  with_lactate %>% rename(first_organ_dysfunction = first_organ_dysfunction_with_lactate)
) %>%
   mutate(site = as.factor(site_name)) #please insert your site name here

# Plot the side-by-side bar chart
first_organ_dysfunction_wi_wo_lactacte %>% 
  filter(!is.na(first_organ_dysfunction),
         !first_organ_dysfunction == "lactate") %>%  
ggplot(aes(x = first_organ_dysfunction, y = percent.total, fill = category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "First Organ Dysfunction", y = "Proportion (%)", 
       title = "Comparison of First Organ Dysfunction With and Without Lactate Criteria") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#If the above graph looks correct, please run this chunk
#note, the 6th row should be NA, which represents patients who only diagnosed with sepsis from lactate 
```{r}
#export table
write_dataframe_to_csv(first_organ_dysfunction_wi_wo_lactacte)

first_organ_dysfunction_wi_wo_lactacte
```

# time gained for part B organ dysfunction diagnosis from lactate inclusion, in patients in which time was gained 
```{r}
# Set maximum number of hours to 96 with an additional bucket for values > 96
max_hours <- 96

# Generate breaks by every hour from 0 to 96 hours, and one extra for greater than 96
breaks <- c(seq(0, max_hours, by = 1), Inf)

# Generate labels for each hourly interval, including one for "greater than 96 hours"
labels <- c(paste0(seq(0, max_hours - 1), "-", seq(1, max_hours), " hours"), "greater than 95 hours")

# Adjust the pipeline to apply hourly breaks up to 96 hours, with an additional bucket for greater than 96
time_gained_with_lactate <- sepsis_case_summary %>%
  select(hospitalization_id, dx_time_gained_with_lactate) %>%
  mutate(
    # Cut the time into hourly intervals up to 96 hours, plus a "greater than 95 hours" bucket
    time_gained_with_lactate = cut(dx_time_gained_with_lactate,
                                   breaks = breaks,
                                   labels = labels,
                                   right = FALSE)  # Left-closed intervals (include lower bound)
  ) %>%
  mutate(time_gained_with_lactate = as.factor(time_gained_with_lactate)) %>%
  group_by(time_gained_with_lactate) %>%
  summarise(count = n_distinct(hospitalization_id)) %>%
  ungroup() %>%
  filter(!is.na(time_gained_with_lactate)) %>% 
  #note the NA at the end of the table is normal, given these are patients who are diagnosed with sepsis only by part B and therefore cant compare time gained to another criteria
  mutate(
    total = sum(count),
    percent.total = round(count / sum(count) * 100, 2)) %>%
  mutate(site = as.factor(site_name))  # Add your site as 'ucmc'

# View the result, up to 20 hours for presentation purposes
target_levels <- paste0(seq(0, 19), "-", seq(1, 20), " hours")

# Filter the dataset to include only the desired factor levels
time_gained_with_lactate_filtered <- time_gained_with_lactate %>%
  filter(time_gained_with_lactate %in% target_levels)  # Check if each value in time_gained_with_lactate is in target_levels

# Create the plot
time_gained_with_lactate_filtered %>% 
  ggplot(aes(x = time_gained_with_lactate, y = percent.total)) +
  geom_bar(stat = "identity") +
 labs(title = "Time Gained for Part B Organ Dysfunction Diagnosis from Lactate Inclusion",
       x = "Time Gained (hours)",
       y = "Proportion (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
# #If the above dsitribution looks correct (majority of time gained should be few hours max), please export your table
#note the last row NA is normal given these are patients who are diagnosed with sepsis only by part B and therefore cant compare time gained 
```{r}
#export table
write_dataframe_to_csv(time_gained_with_lactate)
```

```{r}
clif_patient_cohort %>% count(race_category)
```



# Table 1
# Comparison of organ failure between patients who meet and dont meet part B lactate organ failure 
```{r}
# Mutate the relevant columns and replace NA with 0, non-NA with 1
sepsis_case_summary_organfailures <- sepsis_case_summary %>%
  left_join(clif_hospitalization_cohort %>% 
              select(hospitalization_id, patient_id, discharge_category, age_at_admission, admission_dttm, discharge_dttm) %>%
  mutate(hospitalization_id = as.character(hospitalization_id))) %>% 
  mutate(hospital_los = as.numeric(difftime(discharge_dttm, admission_dttm, units = "days"))) %>% 
  mutate(in_hospital_mortality = ifelse(discharge_category =="Expired", "died", "survived")) %>% 
  left_join(clif_patient_cohort %>% 
              select(patient_id, race_category, ethnicity_category, sex_category) %>%
            mutate(race_category = ifelse(is.na(race_category), "Unknown", race_category))) %>% 
  mutate(
    aki = ifelse(is.na(aki), 0, 1),
    thrombocytopenia = ifelse(is.na(thrombocytopenia), 0, 1),
    invasive_mechanical_ventilation = ifelse(is.na(`invasive mechanical ventilation`), 0, 1),
    vasopressor = ifelse(is.na(vasopressor), 0, 1),
    hyperbilirubinemia = ifelse(is.na(hyperbilirubinemia), 0, 1),
    lactate = ifelse(is.na(lactate), 0, 1)
  ) %>% 
  mutate(inclusion_lactate = 1,
         exclusion_lactate = inclusion_lactate - lactate_only_dx) %>%  #0 for sepsis_exclusion_lactate
  mutate(exclusion_lactate = ifelse(exclusion_lactate == 1, 1, NA)) #1 for sepsis_inclusion_lactate, 0 for sepsis_exclusion_lactate

# Create a function for table 1
lactate_analysis <- function(data) { 
  data %>%
  summarise(
  
    #demographics
    total_n = n_distinct(hospitalization_id),
    age_mean = mean(age_at_admission),
    age_sd = sd(age_at_admission),
    sex_female_n = sum(tolower(sex_category) == "female", na.rm = TRUE),
    sex_female_percent = round(as.double(sex_female_n / total_n * 100), 2),
    sex_male_n = sum(tolower(sex_category) == "male", na.rm = TRUE),
    sex_male_percent = round(as.double(sex_male_n / total_n * 100), 2),
    
    # Race breakdown
    race_black_african_american_n = sum(tolower(race_category) == "black or african american"),
    race_black_african_american_percent = round(race_black_african_american_n/total_n * 100, 2),
    race_white_n = sum(tolower(race_category) == "white"),
    race_white_percent = round(race_white_n/total_n * 100, 2),
    race_asian_n = sum(tolower(race_category) == "asian"),
    race_asian_percent = round(race_asian_n/total_n * 100, 2),
    race_american_indian_alaska_native_n = sum(tolower(race_category) == "american indian or alaska native"),
    race_american_indian_alaska_native_percent = round(race_american_indian_alaska_native_n/total_n * 100, 2),
    race_native_hawaiian_pacific_islander_n = sum(tolower(race_category) == "native hawaiian or other pacific islander"),
    race_native_hawaiian_pacific_islander_percent = round(race_native_hawaiian_pacific_islander_n/total_n * 100, 2),
    race_other_n = sum(tolower(race_category) == "other"),
    race_other_percent = round(race_other_n/total_n * 100, 2),
    race_unknown_n = sum(tolower(race_category) == "unknown"),
    race_unknown_percent = round(race_unknown_n/total_n * 100, 2),
    race_na_n = sum(is.na(race_category)),
    race_na_percent = round(race_na_n/total_n * 100, 2),
    
    # Ethnicity breakdown
    ethnicity_hispanic_n = sum(tolower(ethnicity_category) == "hispanic"),
    ethnicity_hispanic_percent = round(ethnicity_hispanic_n/total_n * 100, 2),
    ethnicity_non_hispanic_n = sum(tolower(ethnicity_category) == "not hispanic"),
    ethnicity_non_hispanic_percent = round(ethnicity_non_hispanic_n/total_n * 100, 2),
    ethnicity_unknown_n = sum(ethnicity_category == "Unknown"),
    ethnicity_unknown_percent = round(ethnicity_unknown_n/total_n * 100, 2),
    
    #organ failures
    aki_n = sum(aki),
    aki_percent = round(aki_n/total_n * 100, 2),
    thrombocytopenia_n = sum(thrombocytopenia),
    thrombocytopenia_percent = round(thrombocytopenia_n/total_n * 100, 2),
    invasive_mechanical_ventilation_n = sum(invasive_mechanical_ventilation),
    invasive_mechanical_ventilation_percent = round(invasive_mechanical_ventilation_n/total_n * 100, 2),
    vasopressor_n = sum(vasopressor),
    vasopressor_percent = round(vasopressor_n/total_n * 100, 2),
    hyperbilirubinemia_n = sum(hyperbilirubinemia),
    hyperbilirubinemia_percent = round(hyperbilirubinemia_n/total_n * 100, 2),
    hospital_los_mean = mean(hospital_los),
    hospital_los_sd = sd(hospital_los),
    died_n = sum(in_hospital_mortality == "died"),
    died_percent = round(died_n/total_n * 100, 2)) 
}

#Run patient cohort exclusion of lactate in sepsis diagnosis
table1_wide_lactate_excuded<- sepsis_case_summary_organfailures %>% 
  group_by(exclusion_lactate) %>%
  lactate_analysis() %>%
  filter(exclusion_lactate == 1) %>% 
  select(-exclusion_lactate) %>% 
  mutate(lactate_inclusion_status = as.factor("lactate_excluded")) %>% 
  select(lactate_inclusion_status, everything()) 

#Run patient cohort inclusion of lactate in sepsis diagnosis
table1_wide_lactate_included <- sepsis_case_summary_organfailures %>% 
  group_by(inclusion_lactate) %>%
  lactate_analysis() %>% 
  select(-inclusion_lactate) %>% 
  mutate(lactate_inclusion_status = as.factor("lactate_included")) %>% 
  select(lactate_inclusion_status, everything()) 

#Run patient cohort lactate only sepsis diagnosis
table1_wide_lactate_only <- sepsis_case_summary_organfailures %>% 
  group_by(lactate_only_dx) %>%
  lactate_analysis() %>% 
  filter(lactate_only_dx == 1) %>%
  select(-lactate_only_dx) %>% 
  mutate(lactate_inclusion_status = as.factor("lactate_only")) %>% 
  select(lactate_inclusion_status, everything()) 

table1_wide <- rbind(table1_wide_lactate_included, table1_wide_lactate_excuded, table1_wide_lactate_only) 

#Reshape 
table1_inclusion_exclusion_lactate_partb <- table1_wide %>%
  pivot_longer(
    cols = -lactate_inclusion_status,  # Pivot all columns except lactate
    names_to = "variable",  # New column for variable names
    values_to = "n") %>% 
  pivot_wider(
    names_from = lactate_inclusion_status,  # Use the 'lactate' column to create new column names
    values_from = n,       # Use the values from the 'n' column
    names_prefix = "" ) %>%  # Prefix for the new columns
  mutate(lactate_included = round(lactate_included, 2),
         lactate_excluded = round(lactate_excluded, 2),
         lactate_only = round(lactate_only, 2)) %>% 
  mutate(site = as.factor(site_name)) #please insert your site name here

print(table1_inclusion_exclusion_lactate_partb)
```
# If the above table looks correct, please export your table into box
```{r}
write_dataframe_to_csv(table1_inclusion_exclusion_lactate_partb)
```

